---
title: "Module 8: Factors in R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "Learn how to create and work with factors in R"
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis::gradethis_setup()
```

## Introduction to Factors in R

Factors are a special data type in R used to represent categorical variables. They are useful for variables that have a fixed and known set of possible values, such as:

- Days of the week
- Months of the year
- Gender
- Blood types
- Rating scales (e.g., "low", "medium", "high")
- Experimental conditions

Factors may look like character vectors, but they are actually integers under the hood, with each integer mapped to a label. This makes them more efficient to store and work with than character vectors.

Key characteristics of factors:
1. They store categorical data efficiently
2. They have predefined levels (possible values)
3. The levels can be ordered (ordinal) or unordered (nominal)
4. They enable proper handling of categorical variables in statistical models and plots

Let's explore how to create, manipulate, and work with factors in R!

## Creating Factors

Factors can be created using the `factor()` function.

### Basic Factor Creation

```{r factor-demo, echo=TRUE}
# Create a character vector
days <- c("Mon", "Tue", "Wed", "Tue", "Mon", "Fri", "Wed")

# Convert to a factor
days_factor <- factor(days)
days_factor

# Check the structure
str(days_factor)

# Get the levels
levels(days_factor)
```

### Specifying Custom Levels

You can specify the levels of a factor, including levels that don't yet exist in the data.

```{r levels-demo, echo=TRUE}
# Create a factor with specified levels
blood_types <- c("AB", "A", "O", "B", "A", "AB")
blood_factor <- factor(blood_types, levels = c("O", "A", "B", "AB"))
blood_factor

# Check the levels
levels(blood_factor)

# What happens with values not in the levels
status <- c("Low", "Medium", "High", "Medium", "Unknown")
status_factor <- factor(status, levels = c("Low", "Medium", "High"))
status_factor  # "Unknown" becomes NA
```

### Ordered Factors

For categorical variables with a natural ordering, you can create ordered factors.

```{r ordered-demo, echo=TRUE}
# Create an ordered factor
sizes <- c("Medium", "Large", "Small", "Small", "Medium", "Large")
sizes_ordered <- factor(sizes, levels = c("Small", "Medium", "Large"), ordered = TRUE)
sizes_ordered

# Check if it's ordered
is.ordered(sizes_ordered)

# Comparison operations work with ordered factors
sizes_ordered[1] > sizes_ordered[3]  # Medium > Small
sizes_ordered[1] < sizes_ordered[2]  # Medium < Large
```

### Exercise 1: Creating Factors

Create the following factors:

1. A basic factor called `weekdays_factor` from this vector: 
   `c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Monday", "Wednesday")`

2. A factor called `ratings_factor` with specified levels from this vector:
   `c("Good", "Excellent", "Poor", "Good", "Fair", "Excellent")`
   Use the levels in ascending order: "Poor", "Fair", "Good", "Excellent"

3. An ordered factor called `education_level` from this vector:
   `c("Bachelor's", "High School", "Master's", "Ph.D.", "High School", "Bachelor's")`
   with levels in this order: "High School", "Bachelor's", "Master's", "Ph.D."

```{r ex1, exercise=TRUE}
# Write your code here

```

```{r ex1-solution}
# 1. Basic factor
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Monday", "Wednesday")
weekdays_factor <- factor(weekdays)

# 2. Factor with specified levels
ratings <- c("Good", "Excellent", "Poor", "Good", "Fair", "Excellent")
ratings_factor <- factor(ratings, levels = c("Poor", "Fair", "Good", "Excellent"))

# 3. Ordered factor
education <- c("Bachelor's", "High School", "Master's", "Ph.D.", "High School", "Bachelor's")
education_level <- factor(education, levels = c("High School", "Bachelor's", "Master's", "Ph.D."), ordered = TRUE)

# Display the factors
weekdays_factor
ratings_factor
education_level
```

```{r ex1-check}
grade_this({
  # Check the student's submitted code using the .user_code object
  user_code <- .user_code
  
  # Check if the code includes all three variable names
  has_weekdays <- grepl("weekdays_factor\\s*<-", user_code)
  has_ratings <- grepl("ratings_factor\\s*<-", user_code)
  has_education <- grepl("education_level\\s*<-", user_code)
  
  # Check for factor() function calls
  uses_factor <- grepl("factor\\(", user_code)
  
  # Check for ordered=TRUE parameter
  uses_ordered <- grepl("ordered\\s*=\\s*TRUE", user_code)
  
  # Check if they created each factor
  if (!has_weekdays || !has_ratings || !has_education) {
    missing_vars <- c()
    if (!has_weekdays) missing_vars <- c(missing_vars, "weekdays_factor")
    if (!has_ratings) missing_vars <- c(missing_vars, "ratings_factor")
    if (!has_education) missing_vars <- c(missing_vars, "education_level")
    fail(paste("Make sure to create all required variables:", paste(missing_vars, collapse=", ")))
  } else if (!uses_factor) {
    fail("Make sure to use the factor() function to create your factors")
  } else if (!uses_ordered && has_education) {
    fail("Make sure education_level is an ordered factor (use ordered=TRUE)")
  } else {
    pass("Great job! You've created all the required factors correctly.")
  }
})
```
```

## Working with Factor Levels

Once a factor is created, you can work with its levels in various ways:

### Accessing and Modifying Levels

```{r modify-levels-demo, echo=TRUE}
# Create a factor
colors <- factor(c("Red", "Blue", "Green", "Red", "Yellow", "Blue"))

# Get levels
levels(colors)

# Change level names
levels(colors) <- c("Rouge", "Bleu", "Vert", "Jaune")
colors

# Change specific level
levels(colors)[levels(colors) == "Rouge"] <- "Red"
colors
```

### Recoding Factor Levels

```{r recode-demo, echo=TRUE}
# Create a factor
responses <- factor(c(1, 2, 3, 3, 1, 2, 3, 2, 1))

# Recode with descriptive labels
levels(responses) <- c("Disagree", "Neutral", "Agree")
responses
```

### Combining Levels

```{r combine-demo, echo=TRUE}
# Create a factor
income <- factor(c("Low", "Medium", "High", "Medium", "Low", "Very High", "Medium"))

# Combine "High" and "Very High" into one level
levels(income)[levels(income) %in% c("High", "Very High")] <- "High+"
income
```

### Exercise 2: Working with Factor Levels

Start with this factor:

```
survey_responses <- factor(c("Strongly Agree", "Agree", "Neutral", "Disagree", 
                            "Strongly Disagree", "Neutral", "Agree", "Strongly Agree",
                            "Disagree", "Neutral"))
```

Complete the following tasks:
1. Display the levels of the factor
2. Change the levels to be in logical order: "Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree"
3. Simplify the factor by combining "Strongly Agree" and "Agree" into "Positive", and "Strongly Disagree" and "Disagree" into "Negative", leaving "Neutral" as is
4. Count the frequency of each level in the simplified factor

```{r ex2, exercise=TRUE}
# Create the survey_responses factor
survey_responses <- factor(c("Strongly Agree", "Agree", "Neutral", "Disagree", 
                            "Strongly Disagree", "Neutral", "Agree", "Strongly Agree",
                            "Disagree", "Neutral"))

# Write your code here

```

```{r ex2-solution}
# Create the survey_responses factor
survey_responses <- factor(c("Strongly Agree", "Agree", "Neutral", "Disagree", 
                            "Strongly Disagree", "Neutral", "Agree", "Strongly Agree",
                            "Disagree", "Neutral"))

# 1. Display the levels
original_levels <- levels(survey_responses)
original_levels

# 2. Change the order of levels
survey_responses <- factor(survey_responses, 
                          levels = c("Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree"))

# Display the reordered factor
survey_responses

# 3. Simplify by combining levels
simplified_responses <- survey_responses  # Make a copy
levels(simplified_responses) <- c("Negative", "Negative", "Neutral", "Positive", "Positive")

# Display the simplified factor
simplified_responses

# 4. Count frequency of each level
response_counts <- table(simplified_responses)
response_counts
```

```{r ex2-check}
grade_this({
  # Check the student's submitted code
  user_code <- .user_code
  
  # Check for required variable assignments
  has_survey <- grepl("survey_responses\\s*<-", user_code)
  has_simplified <- grepl("simplified_responses\\s*<-", user_code)
  has_counts <- grepl("response_counts\\s*<-", user_code)
  
  # Check for required functions
  uses_levels <- grepl("levels\\(", user_code)
  uses_factor <- grepl("factor\\(", user_code)
  uses_table <- grepl("table\\(", user_code)
  
  # Check if all required variables are created
  if (!has_survey || !has_simplified || !has_counts) {
    missing_vars <- c()
    if (!has_survey) missing_vars <- c(missing_vars, "survey_responses (reordered)")
    if (!has_simplified) missing_vars <- c(missing_vars, "simplified_responses")
    if (!has_counts) missing_vars <- c(missing_vars, "response_counts")
    
    fail(paste("Make sure to create:", paste(missing_vars, collapse=", ")))
  } else if (!uses_levels) {
    fail("Make sure to use the levels() function to modify factor levels")
  } else if (!uses_table) {
    fail("Make sure to use the table() function to count frequencies")
  } else {
    pass("Excellent! You've successfully completed all the factor level operations.")
  }
})
```

## Converting Between Factors and Other Types

You can convert factors to and from other data types:

### Converting Factors to Other Types

```{r convert-from-demo, echo=TRUE}
# Create a factor
ratings <- factor(c("Low", "Medium", "High", "Medium", "Low", "High"))

# Convert factor to character
ratings_char <- as.character(ratings)
ratings_char

# Convert factor to numeric (returns the internal integer codes)
# Note: This is usually not what you want
ratings_num <- as.numeric(ratings)
ratings_num

# To convert to meaningful numbers, convert to character first
# Then to numeric (only if the levels are numeric strings)
num_factor <- factor(c("10", "20", "30", "20", "10"))
as.numeric(as.character(num_factor))
```

### Converting Other Types to Factors

```{r convert-to-demo, echo=TRUE}
# Character vector to factor
days <- c("Mon", "Tue", "Wed", "Thu", "Fri")
days_factor <- as.factor(days)
days_factor

# Numeric vector to factor
codes <- c(1, 2, 3, 2, 1, 3, 2)
codes_factor <- as.factor(codes)
codes_factor
```

### Exercise 3: Converting Between Factors and Other Types

Complete the following tasks:
1. Create a factor `temp_factor` from this character vector:
   `c("Cold", "Hot", "Cold", "Warm", "Hot", "Cold", "Warm")`

2. Convert `temp_factor` to a character vector called `temp_char`

3. Create a factor `score_factor` from this numeric vector:
   `c(1, 2, 3, 1, 3, 2, 1, 2)`

4. Assign descriptive labels to `score_factor`: 1="Low", 2="Medium", 3="High"
   Store the result as `labeled_scores`

5. Create a factor from this character vector of numeric strings:
   `c("10", "20", "30", "20", "10", "30")`
   Then convert it to an actual numeric vector and store as `numeric_values`

```{r ex3, exercise=TRUE}
# Write your code here

```

```{r ex3-solution}
# 1. Create temperature factor
temp_factor <- factor(c("Cold", "Hot", "Cold", "Warm", "Hot", "Cold", "Warm"))

# 2. Convert to character
temp_char <- as.character(temp_factor)

# 3. Create score factor from numeric vector
score_factor <- factor(c(1, 2, 3, 1, 3, 2, 1, 2))

# 4. Assign descriptive labels
labeled_scores <- score_factor
levels(labeled_scores) <- c("Low", "Medium", "High")

# 5. Convert numeric strings
numeric_strings <- factor(c("10", "20", "30", "20", "10", "30"))
numeric_values <- as.numeric(as.character(numeric_strings))

# Display results
temp_factor
temp_char
score_factor
labeled_scores
numeric_values
```

```{r ex3-check}
grade_this({
  # Check the student's submitted code
  user_code <- .user_code
  
  # Check for required variable assignments
  has_temp_factor <- grepl("temp_factor\\s*<-", user_code)
  has_temp_char <- grepl("temp_char\\s*<-", user_code)
  has_score_factor <- grepl("score_factor\\s*<-", user_code)
  has_labeled_scores <- grepl("labeled_scores\\s*<-", user_code)
  has_numeric_values <- grepl("numeric_values\\s*<-", user_code)
  
  # Check for required functions
  uses_factor <- grepl("factor\\(", user_code)
  uses_as_character <- grepl("as\\.character\\(", user_code)
  uses_as_numeric <- grepl("as\\.numeric\\(", user_code)
  uses_levels <- grepl("levels\\(", user_code)
  
  # Check if all required variables are created
  if (!has_temp_factor || !has_temp_char || !has_score_factor || !has_labeled_scores || !has_numeric_values) {
    missing_vars <- c()
    if (!has_temp_factor) missing_vars <- c(missing_vars, "temp_factor")
    if (!has_temp_char) missing_vars <- c(missing_vars, "temp_char")
    if (!has_score_factor) missing_vars <- c(missing_vars, "score_factor")
    if (!has_labeled_scores) missing_vars <- c(missing_vars, "labeled_scores")
    if (!has_numeric_values) missing_vars <- c(missing_vars, "numeric_values")
    
    fail(paste("Make sure to create all five variables:", paste(missing_vars, collapse=", ")))
  } else if (!uses_factor) {
    fail("Make sure to use the factor() function to create factors")
  } else if (!uses_as_character) {
    fail("Make sure to use as.character() to convert from factor to character")
  } else if (!uses_as_numeric) {
    fail("Make sure to use as.numeric() to convert to numeric values")
  } else if (!uses_levels && has_labeled_scores) {
    fail("Make sure to use levels() to assign descriptive labels to score_factor")
  } else {
    pass("Great job! All conversions between factors and other types are correct.")
  }
})
```


## Factors in Data Analysis

Factors are particularly useful in data analysis for:

### Tabulation and Summarization

```{r tabulation-demo, echo=TRUE}
# Create a factor
gender <- factor(c("Male", "Female", "Female", "Male", "Male", "Female", "Female", "Male"))

# Tabulate frequencies
table(gender)

# Cross-tabulation
treatment <- factor(c("A", "A", "B", "B", "A", "B", "A", "B"))
table(gender, treatment)

# Proportions
prop.table(table(gender))
prop.table(table(gender, treatment))  # Joint proportions
prop.table(table(gender, treatment), 1)  # Row proportions
prop.table(table(gender, treatment), 2)  # Column proportions
```

### Plotting with Factors

```{r plot-demo, echo=TRUE}
# Create some example data
education <- factor(c("High School", "Bachelor's", "Master's", "Ph.D.", 
                      "High School", "Bachelor's", "Master's", "High School",
                      "Bachelor's", "Master's", "Ph.D.", "High School"),
                   levels = c("High School", "Bachelor's", "Master's", "Ph.D."),
                   ordered = TRUE)

salary <- c(45000, 65000, 80000, 100000, 42000, 63000, 81000, 47000, 66000, 79000, 105000, 44000)

# Bar plot of education frequencies
barplot(table(education), col = "lightblue", main = "Education Levels")

# Boxplot of salary by education
plot(education, salary, col = "lightgreen", main = "Salary by Education Level",
     xlab = "Education", ylab = "Salary")
```

### Exercise 4: Working with Factors in Data Analysis

Use the following data about customer satisfaction ratings at a restaurant:

```
ratings <- factor(c("Excellent", "Good", "Fair", "Poor", "Good", "Excellent", 
                   "Good", "Fair", "Excellent", "Good", "Poor", "Fair", 
                   "Good", "Excellent", "Excellent", "Good", "Fair", "Poor"),
                  levels = c("Poor", "Fair", "Good", "Excellent"),
                  ordered = TRUE)

expenditure <- c(45, 32, 18, 15, 30, 50, 27, 20, 55, 34, 12, 22, 28, 48, 52, 35, 21, 14)
```

Complete the following tasks:
1. Create a frequency table of the ratings and store as `rating_freq`
2. Calculate the percentage of each rating and store as `rating_pct`
3. Create a data frame called `rating_data` with the ratings and expenditure
4. Calculate the average expenditure for each rating level and store as `avg_by_rating`
5. Create a barplot of the rating frequencies

```{r ex4, exercise=TRUE}
# Create the ratings factor and expenditure vector
ratings <- factor(c("Excellent", "Good", "Fair", "Poor", "Good", "Excellent", 
                   "Good", "Fair", "Excellent", "Good", "Poor", "Fair", 
                   "Good", "Excellent", "Excellent", "Good", "Fair", "Poor"),
                  levels = c("Poor", "Fair", "Good", "Excellent"),
                  ordered = TRUE)

expenditure <- c(45, 32, 18, 15, 30, 50, 27, 20, 55, 34, 12, 22, 28, 48, 52, 35, 21, 14)

# Write your code here

```

```{r ex4-solution}
# Create the ratings factor and expenditure vector
ratings <- factor(c("Excellent", "Good", "Fair", "Poor", "Good", "Excellent", 
                   "Good", "Fair", "Excellent", "Good", "Poor", "Fair", 
                   "Good", "Excellent", "Excellent", "Good", "Fair", "Poor"),
                  levels = c("Poor", "Fair", "Good", "Excellent"),
                  ordered = TRUE)

expenditure <- c(45, 32, 18, 15, 30, 50, 27, 20, 55, 34, 12, 22, 28, 48, 52, 35, 21, 14)

# 1. Create a frequency table
rating_freq <- table(ratings)

# 2. Calculate percentages
rating_pct <- prop.table(rating_freq) * 100

# 3. Create a data frame
rating_data <- data.frame(Rating = ratings, Expenditure = expenditure)

# 4. Calculate average expenditure by rating
avg_by_rating <- aggregate(Expenditure ~ Rating, data = rating_data, FUN = mean)

# 5. Create a barplot
barplot(rating_freq, col = c("red", "orange", "green", "blue"), 
        main = "Distribution of Customer Ratings",
        xlab = "Rating", ylab = "Frequency")

# Display results
rating_freq
rating_pct
rating_data
avg_by_rating
```

```{r ex4-check}
grade_this({
  # Check the student's submitted code
  user_code <- .user_code
  
  # Check for required variable assignments
  has_rating_freq <- grepl("rating_freq\\s*<-", user_code)
  has_rating_pct <- grepl("rating_pct\\s*<-", user_code)
  has_rating_data <- grepl("rating_data\\s*<-", user_code)
  has_avg_by_rating <- grepl("avg_by_rating\\s*<-", user_code)
  
  # Check for required functions
  uses_table <- grepl("table\\(", user_code)
  uses_prop_table <- grepl("prop\\.table\\(", user_code)
  uses_data_frame <- grepl("data\\.frame\\(", user_code)
  uses_aggregate <- grepl("aggregate\\(", user_code)
  uses_barplot <- grepl("barplot\\(", user_code)
  
  # Check if all required variables are created
  if (!has_rating_freq || !has_rating_pct || !has_rating_data || !has_avg_by_rating) {
    missing_vars <- c()
    if (!has_rating_freq) missing_vars <- c(missing_vars, "rating_freq")
    if (!has_rating_pct) missing_vars <- c(missing_vars, "rating_pct")
    if (!has_rating_data) missing_vars <- c(missing_vars, "rating_data")
    if (!has_avg_by_rating) missing_vars <- c(missing_vars, "avg_by_rating")
    
    fail(paste("Make sure to create all variables:", paste(missing_vars, collapse=", ")))
  } else if (!uses_table) {
    fail("Make sure to use the table() function to create a frequency table")
  } else if (!uses_prop_table) {
    fail("Make sure to use prop.table() to calculate percentages")
  } else if (!uses_data_frame) {
    fail("Make sure to use data.frame() to create a data frame with ratings and expenditures")
  } else if (!uses_aggregate) {
    fail("Make sure to use aggregate() to calculate average expenditure by rating")
  } else if (!uses_barplot) {
    fail("Make sure to create a barplot of the rating frequencies")
  } else {
    pass("Excellent! You've successfully completed all the factor data analysis tasks.")
  }
})
```



## Factors in Statistical Modeling

Factors are particularly important for statistical modeling in R.

### Model Building with Factors

In statistical models, factors are automatically handled differently from numeric variables:
- Factors with two levels are typically converted to 0/1 dummy variables
- Factors with more than two levels are converted to a set of dummy variables

```{r model-demo, echo=TRUE}
# Create a simple dataset
set.seed(123)  # For reproducibility
treatment <- factor(rep(c("Control", "Drug A", "Drug B"), each = 10))
age <- sample(30:70, 30, replace = TRUE)
response <- rnorm(30, mean = c(5, 7, 9)[as.numeric(treatment)], sd = 1.5) + age/10

# Build a linear model
model <- lm(response ~ treatment + age)

# View the model summary
summary(model)
```

### Exercise 5: Factors in Statistical Modeling

Use this simulated dataset about plant growth under different conditions:

```
# Generate data
set.seed(456)
fertilizer <- factor(rep(c("None", "Type A", "Type B"), each = 15))
sunshine <- factor(rep(c("Low", "Medium", "High"), times = 15))
water <- sample(5:15, 45, replace = TRUE)
growth <- 2 + 0.5 * water + 
          rnorm(45, mean = c(0, 2, 3)[as.numeric(fertilizer)], sd = 1) +
          rnorm(45, mean = c(0, 1, 2)[as.numeric(sunshine)], sd = 0.5)
```

Complete the following tasks:
1. Create a data frame called `plant_data` with all variables
2. Check the structure of the data frame to verify factors are correctly recognized
3. Calculate the average growth by fertilizer type and save as `growth_by_fertilizer`
4. Calculate the average growth by sunshine level and save as `growth_by_sunshine`
5. Build a linear model that predicts `growth` based on `fertilizer`, `sunshine`, and `water`, and save as `plant_model`

```{r ex5, exercise=TRUE}
# Generate data
set.seed(456)
fertilizer <- factor(rep(c("None", "Type A", "Type B"), each = 15))
sunshine <- factor(rep(c("Low", "Medium", "High"), times = 15))
water <- sample(5:15, 45, replace = TRUE)
growth <- 2 + 0.5 * water + 
          rnorm(45, mean = c(0, 2, 3)[as.numeric(fertilizer)], sd = 1) +
          rnorm(45, mean = c(0, 1, 2)[as.numeric(sunshine)], sd = 0.5)

# Write your code here

```

```{r ex5-solution}
# Generate data
set.seed(456)
fertilizer <- factor(rep(c("None", "Type A", "Type B"), each = 15))
sunshine <- factor(rep(c("Low", "Medium", "High"), times = 15))
water <- sample(5:15, 45, replace = TRUE)
growth <- 2 + 0.5 * water + 
          rnorm(45, mean = c(0, 2, 3)[as.numeric(fertilizer)], sd = 1) +
          rnorm(45, mean = c(0, 1, 2)[as.numeric(sunshine)], sd = 0.5)

# 1. Create a data frame
plant_data <- data.frame(
  Fertilizer = fertilizer,
  Sunshine = sunshine,
  Water = water,
  Growth = growth
)

# 2. Check the structure
str(plant_data)

# 3. Calculate average growth by fertilizer type
growth_by_fertilizer <- aggregate(Growth ~ Fertilizer, data = plant_data, FUN = mean)

# 4. Calculate average growth by sunshine level
growth_by_sunshine <- aggregate(Growth ~ Sunshine, data = plant_data, FUN = mean)

# 5. Build a linear model
plant_model <- lm(Growth ~ Fertilizer + Sunshine + Water, data = plant_data)

# Display results
plant_data
growth_by_fertilizer
growth_by_sunshine
summary(plant_model)
```

```{r ex5-check}
grade_this({
  # Check the student's submitted code
  user_code <- .user_code
  
  # Check for required variable assignments
  has_plant_data <- grepl("plant_data\\s*<-", user_code)
  has_growth_by_fertilizer <- grepl("growth_by_fertilizer\\s*<-", user_code)
  has_growth_by_sunshine <- grepl("growth_by_sunshine\\s*<-", user_code)
  has_plant_model <- grepl("plant_model\\s*<-", user_code)
  
  # Check for required functions
  uses_data_frame <- grepl("data\\.frame\\(", user_code)
  uses_str <- grepl("str\\(", user_code)
  uses_aggregate <- grepl("aggregate\\(", user_code)
  uses_lm <- grepl("lm\\(", user_code)
  
  # Check if all required variables are created
  if (!has_plant_data || !has_growth_by_fertilizer || !has_growth_by_sunshine || !has_plant_model) {
    missing_vars <- c()
    if (!has_plant_data) missing_vars <- c(missing_vars, "plant_data")
    if (!has_growth_by_fertilizer) missing_vars <- c(missing_vars, "growth_by_fertilizer")
    if (!has_growth_by_sunshine) missing_vars <- c(missing_vars, "growth_by_sunshine")
    if (!has_plant_model) missing_vars <- c(missing_vars, "plant_model")
    
    fail(paste("Make sure to create all variables:", paste(missing_vars, collapse=", ")))
  } else if (!uses_data_frame) {
    fail("Make sure to use data.frame() to create a data frame with all variables")
  } else if (!uses_str) {
    fail("Make sure to check the structure of the data frame using str()")
  } else if (!uses_aggregate) {
    fail("Make sure to use aggregate() to calculate average growth by fertilizer and sunshine")
  } else if (!uses_lm) {
    fail("Make sure to use lm() to build a linear model predicting growth")
  } else {
    pass("Excellent! You've successfully completed all the factor modeling tasks.")
  }
})
```

## Summary

In this module, you've learned about factors in R, including:

1. Creating factors using the `factor()` function with specified levels and ordering
2. Working with factor levels by accessing, modifying, and combining them
3. Converting between factors and other data types such as character and numeric
4. Using factors for data analysis through tabulation, summarization, and visualization
5. Applying factors in statistical modeling to represent categorical variables

Factors are essential for proper handling of categorical data in R. They ensure that categories are correctly represented in analyses, visualizations, and statistical models, which is crucial for accurate data interpretation and decision-making.

This concludes our tutorial on basic data structures in R. You've now learned about the environment, variables, data types, vectors, matrices, lists, data frames, and factors—all the fundamental building blocks you need for effective data analysis in R!
