---
title: "Module 4: Vectors in R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "Learn how to create and manipulate vectors in R"
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis::gradethis_setup()
```

## Introduction to Vectors in R

Vectors are one of the most fundamental data structures in R. A vector is a one-dimensional array that can hold multiple values of the same data type. Almost everything in R is built upon or derived from vectors.

Key characteristics of vectors:
1. All elements must have the same data type (numeric, character, logical, etc.)
2. Vectors are ordered, meaning each element has a specific position
3. Elements can be accessed by their index using square brackets `[]`

Vectors are extensively used in R for:
- Storing related values (e.g., a series of measurements)
- Performing operations on multiple values at once (vectorized operations)
- Creating more complex data structures like matrices and data frames

Let's explore how to create, manipulate, and work with vectors in R!

## Creating Vectors

There are several ways to create vectors in R:

### Using the combine function `c()`

The most common way to create a vector is using the `c()` function, which combines values into a vector.

```{r c-demo, echo=TRUE}
# Numeric vector
numbers <- c(10, 20, 30, 40, 50)
numbers

# Character vector
fruits <- c("apple", "banana", "orange", "grape")
fruits

# Logical vector
flags <- c(TRUE, FALSE, TRUE, TRUE)
flags
```

### Using sequence functions

R provides several functions to create vectors with sequential values:

- `seq()` or `:` operator for creating sequences
- `rep()` for repeating values

```{r seq-demo, echo=TRUE}
# Using the colon operator for integer sequences
seq1 <- 1:10
seq1

# Using seq() for more control
seq2 <- seq(from = 0, to = 1, by = 0.2)
seq2

seq3 <- seq(from = 0, to = 10, length.out = 5)
seq3

# Using rep() for repetition
rep1 <- rep(5, times = 4)
rep1

rep2 <- rep(c(1, 2, 3), each = 2)
rep2

rep3 <- rep(c(1, 2, 3), times = 2)
rep3
```

### Exercise 1: Creating Vectors

Create the following vectors:
1. A numeric vector named `temperatures` containing the values 68, 72, 75, 69, 71
2. A character vector named `days` containing the days of the week (Monday through Sunday)
3. A sequence vector named `countdown` with values from 10 down to 1

```{r ex1, exercise=TRUE}
# Write your code here

```

```{r ex1-solution}
# Numeric vector
temperatures <- c(68, 72, 75, 69, 71)

# Character vector
days <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

# Sequence vector (counting down)
countdown <- 10:1

# Display vectors
temperatures
days
countdown
```

```{r ex1-check}
grade_this({
  # Check for pattern matches in the code
  has_temperatures <- grepl("temperatures\\s*<-\\s*c\\s*\\(\\s*68\\s*,\\s*72\\s*,\\s*75\\s*,\\s*69\\s*,\\s*71\\s*\\)", .user_code)
  
  has_days <- grepl('days\\s*<-\\s*c\\s*\\(\\s*["\'](Monday|monday)["\'](\\s*,\\s*["\'].*["\']+\\s*)+\\)', .user_code) && 
              grepl('["\'](Sunday|sunday)["\']', .user_code) &&
              grepl('["\'](Saturday|saturday)["\']', .user_code)
  
  has_countdown <- grepl("countdown\\s*<-\\s*(10\\s*:\\s*1|seq\\s*\\(\\s*10\\s*,\\s*1\\s*)|rev\\s*\\(\\s*1\\s*:\\s*10\\s*\\))", .user_code)
  
  # Check if the outputs match expected values
  if (exists("temperatures", envir = environment()) && exists("days", envir = environment()) && exists("countdown", envir = environment())) {
    temp_matches <- is.numeric(temperatures) && length(temperatures) == 5 && 
                   all(temperatures == c(68, 72, 75, 69, 71))
    
    days_matches <- is.character(days) && length(days) == 7 && 
                    any(grepl("monday", tolower(days))) && 
                    any(grepl("sunday", tolower(days)))
    
    countdown_matches <- is.numeric(countdown) && length(countdown) == 10 && 
                          countdown[1] == 10 && countdown[10] == 1
    
    if (temp_matches && days_matches && countdown_matches) {
      pass("Great job! You've successfully created all three vectors.")
    } else if (!temp_matches) {
      fail("Your temperatures vector doesn't match the expected values (68, 72, 75, 69, 71).")
    } else if (!days_matches) {
      fail("Your days vector doesn't contain all seven days of the week (Monday through Sunday).")
    } else if (!countdown_matches) {
      fail("Your countdown vector should contain numbers from 10 down to 1.")
    } else {
      fail("Check that your vectors match the required specifications.")
    }
  } else if (has_temperatures && has_days && has_countdown) {
    pass("Great job! You've successfully created all three vectors.")
  } else if (!has_temperatures) {
    fail("Make sure to create a numeric vector named 'temperatures' with values 68, 72, 75, 69, 71.")
  } else if (!has_days) {
    fail("Make sure to create a character vector named 'days' with all seven days of the week (Monday through Sunday).")
  } else if (!has_countdown) {
    fail("Make sure to create a vector named 'countdown' with values from 10 down to 1 (10, 9, 8, ..., 1).")
  } else {
    fail("Make sure to create all three vectors with the correct names and values.")
  }
})
```

## Vector Operations

One of the most powerful features of R is its ability to perform operations on entire vectors at once (vectorized operations). This makes code concise and efficient.

### Arithmetic Operations

When you perform arithmetic operations on vectors, R applies the operation element-wise.

```{r vector-arith-demo, echo=TRUE}
# Create vectors
a <- c(1, 2, 3, 4, 5)
b <- c(10, 20, 30, 40, 50)

# Element-wise operations
a + b
a * b
a / b
a ^ 2  # Square each element

# Operations with a single value
a + 10
a * 2
```

### Vector Recycling

When operating on vectors of different lengths, R will "recycle" (repeat) the shorter vector to match the length of the longer one.

```{r recycling-demo, echo=TRUE}
# Vector recycling
x <- c(1, 2, 3, 4, 5)
y <- c(10, 20)

# R will recycle y to c(10, 20, 10, 20, 10)
x + y

# With a warning if not a multiple
z <- c(10, 20, 30)
x + z  # Warning: length of x is not a multiple of length of z
```

### Exercise 2: Vector Operations

Create two numeric vectors:
- `prices` with values 10.99, 24.50, 5.75, 18.20
- `quantities` with values 2, 1, 5, 3

Then calculate and store:
1. The `total_cost` for each item (price × quantity)
2. A `discounted_prices` vector with a 15% discount applied to each price
3. The `item_discount` amount saved for each item (original cost - discounted cost)

```{r ex2, exercise=TRUE}
# Write your code here

```

```{r ex2-solution}
# Create vectors
prices <- c(10.99, 24.50, 5.75, 18.20)
quantities <- c(2, 1, 5, 3)

# Calculate totals
total_cost <- prices * quantities

# Apply discount
discounted_prices <- prices * 0.85  # 15% discount means keeping 85%

# Calculate savings per item
item_discount <- (prices * quantities) - (discounted_prices * quantities)

# Display results
prices
quantities
total_cost
discounted_prices
item_discount
```

```{r ex2-check}
grade_this({
  # Check for pattern matches in the code
  has_prices <- grepl("prices\\s*<-\\s*c\\s*\\(\\s*10\\.99\\s*,\\s*24\\.50\\s*,\\s*5\\.75\\s*,\\s*18\\.20\\s*\\)", .user_code)
  
  has_quantities <- grepl("quantities\\s*<-\\s*c\\s*\\(\\s*2\\s*,\\s*1\\s*,\\s*5\\s*,\\s*3\\s*\\)", .user_code)
  
  has_total_cost <- grepl("total_cost\\s*<-\\s*prices\\s*\\*\\s*quantities", .user_code)
  
  has_discount <- grepl("discounted_prices\\s*<-\\s*prices\\s*\\*\\s*(0\\.85|1\\s*-\\s*0\\.15)|prices\\s*-\\s*prices\\s*\\*\\s*0\\.15", .user_code)
  
  has_item_discount <- grepl("item_discount\\s*<-\\s*\\(prices\\s*\\*\\s*quantities\\)\\s*-\\s*\\(discounted_prices\\s*\\*\\s*quantities\\)", .user_code)
  
  if (exists("prices", envir = environment()) && 
      exists("quantities", envir = environment()) && 
      exists("total_cost", envir = environment()) && 
      exists("discounted_prices", envir = environment()) && 
      exists("item_discount", envir = environment())) {
    
    prices_match <- all(abs(prices - c(10.99, 24.50, 5.75, 18.20)) < 0.001)
    quantities_match <- all(quantities == c(2, 1, 5, 3))
    total_cost_match <- all(abs(total_cost - (prices * quantities)) < 0.001)
    discounted_match <- all(abs(discounted_prices - (prices * 0.85)) < 0.001)
    discount_match <- all(abs(item_discount - ((prices * quantities) - (discounted_prices * quantities))) < 0.001)
    
    if (prices_match && quantities_match && total_cost_match && discounted_match && discount_match) {
      pass("Excellent! All your vector calculations are correct.")
    } else {
      if (!prices_match) {
        fail("Your prices vector should contain 10.99, 24.50, 5.75, 18.20.")
      } else if (!quantities_match) {
        fail("Your quantities vector should contain 2, 1, 5, 3.")
      } else if (!total_cost_match) {
        fail("Check your total_cost calculation. It should be prices * quantities.")
      } else if (!discounted_match) {
        fail("Check your discounted_prices calculation. It should be prices * 0.85 (applying 15% discount).")
      } else if (!discount_match) {
        fail("Check your item_discount calculation. It should be the difference between original and discounted total costs.")
      }
    }
  } else if (has_prices && has_quantities && has_total_cost && has_discount && has_item_discount) {
    pass("Excellent! All your vector calculations are correct.")
  } else if (!has_prices) {
    fail("Make sure to create the prices vector with values 10.99, 24.50, 5.75, 18.20.")
  } else if (!has_quantities) {
    fail("Make sure to create the quantities vector with values 2, 1, 5, 3.")
  } else if (!has_total_cost) {
    fail("Make sure to calculate total_cost as prices * quantities.")
  } else if (!has_discount) {
    fail("Make sure to calculate discounted_prices by applying a 15% discount to prices.")
  } else if (!has_item_discount) {
    fail("Make sure to calculate item_discount as the difference between original and discounted costs.")
  } else {
    fail("Make sure to create all five vectors with the correct calculations.")
  }
})
```

## Vector Indexing and Subsetting

You can access elements of a vector using indexing, which allows you to work with specific elements or subsets of a vector.

### Indexing with Positions

In R, indices start at 1 (not 0 as in many other programming languages).

```{r indexing-demo, echo=TRUE}
# Create a vector
x <- c(10, 20, 30, 40, 50)

# Access a single element
x[1]  # First element (10)
x[3]  # Third element (30)

# Access multiple elements
x[c(1, 3, 5)]  # First, third, and fifth elements

# Access a range of elements
x[2:4]  # Elements 2 through 4
```

### Negative Indexing

Using negative indices excludes those elements.

```{r neg-indexing-demo, echo=TRUE}
# Create a vector
x <- c(10, 20, 30, 40, 50)

# Exclude elements
x[-1]  # All elements except the first
x[-c(1, 3)]  # All elements except the first and third
```

### Logical Indexing

You can use a logical vector to select elements that correspond to TRUE.

```{r logical-indexing-demo, echo=TRUE}
# Create a vector
x <- c(10, 20, 30, 40, 50)

# Create a logical condition
condition <- x > 25

# See the logical vector
condition

# Use logical indexing
x[condition]  # Elements greater than 25

# Directly in the index
x[x > 25]  # Same result
x[x >= 20 & x <= 40]  # Elements between 20 and 40 (inclusive)
```

### Exercise 3: Vector Indexing

Using this vector of monthly sales data:
```
monthly_sales <- c(15200, 18500, 22100, 19800, 24500, 26300, 22800, 23900, 25600, 21400, 18900, 20300)
```

Complete the following tasks:
1. Extract the sales figures for the first quarter (first 3 months)
2. Extract the sales figures for the last quarter (last 3 months)
3. Find all months with sales greater than 22000
4. Find the sales figures for months 5, 7, and 10

```{r ex3, exercise=TRUE}
# Create the monthly sales vector
monthly_sales <- c(15200, 18500, 22100, 19800, 24500, 26300, 22800, 23900, 25600, 21400, 18900, 20300)

# Write your code here

```

```{r ex3-solution}
# Create the monthly sales vector
monthly_sales <- c(15200, 18500, 22100, 19800, 24500, 26300, 22800, 23900, 25600, 21400, 18900, 20300)

# First quarter
q1_sales <- monthly_sales[1:3]
q1_sales

# Last quarter
q4_sales <- monthly_sales[10:12]
q4_sales

# Sales greater than 22000
high_sales <- monthly_sales[monthly_sales > 22000]
high_sales

# Specific months
selected_months <- monthly_sales[c(5, 7, 10)]
selected_months
```

```{r ex3-check}
grade_this({
  # Check for pattern matches in the code
  has_q1 <- grepl("q1_sales\\s*<-\\s*monthly_sales\\s*\\[\\s*1\\s*:\\s*3\\s*\\]", .user_code)
  
  has_q4 <- grepl("q4_sales\\s*<-\\s*monthly_sales\\s*\\[\\s*10\\s*:\\s*12\\s*\\]", .user_code)
  
  has_high <- grepl("high_sales\\s*<-\\s*monthly_sales\\s*\\[\\s*monthly_sales\\s*>\\s*22000\\s*\\]", .user_code)
  
  has_selected <- grepl("selected_months\\s*<-\\s*monthly_sales\\s*\\[\\s*c\\s*\\(\\s*5\\s*,\\s*7\\s*,\\s*10\\s*\\)\\s*\\]", .user_code)
  
  if (exists("q1_sales", envir = environment()) && 
      exists("q4_sales", envir = environment()) && 
      exists("high_sales", envir = environment()) && 
      exists("selected_months", envir = environment())) {
    
    q1_match <- all(q1_sales == monthly_sales[1:3])
    q4_match <- all(q4_sales == monthly_sales[10:12])
    high_match <- all(high_sales == monthly_sales[monthly_sales > 22000])
    selected_match <- all(selected_months == monthly_sales[c(5, 7, 10)])
    
    if (q1_match && q4_match && high_match && selected_match) {
      pass("Perfect! All your vector indexing operations are correct.")
    } else {
      if (!q1_match) {
        fail("Your q1_sales should contain the first three elements of monthly_sales.")
      } else if (!q4_match) {
        fail("Your q4_sales should contain the last three elements of monthly_sales.")
      } else if (!high_match) {
        fail("Your high_sales should contain all elements greater than 22000.")
      } else if (!selected_match) {
        fail("Your selected_months should contain elements 5, 7, and 10 from monthly_sales.")
      }
    }
  } else if (has_q1 && has_q4 && has_high && has_selected) {
    pass("Perfect! All your vector indexing operations are correct.")
  } else if (!has_q1) {
    fail("Make sure to extract the first quarter sales as q1_sales <- monthly_sales[1:3].")
  } else if (!has_q4) {
    fail("Make sure to extract the last quarter sales as q4_sales <- monthly_sales[10:12].")
  } else if (!has_high) {
    fail("Make sure to find months with sales greater than 22000 as high_sales <- monthly_sales[monthly_sales > 22000].")
  } else if (!has_selected) {
    fail("Make sure to extract specific months as selected_months <- monthly_sales[c(5, 7, 10)].")
  } else {
    fail("Make sure to complete all four vector indexing operations.")
  }
})
```

## Vector Functions

R provides many built-in functions for working with vectors:

### Statistical Functions

```{r stat-func-demo, echo=TRUE}
# Create a numeric vector
x <- c(5, 10, 15, 20, 25, 30)

# Basic statistics
sum(x)      # Sum of all elements
mean(x)     # Mean (average)
median(x)   # Median
min(x)      # Minimum value
max(x)      # Maximum value
range(x)    # Range (minimum and maximum)
sd(x)       # Standard deviation
var(x)      # Variance
```

### Other Useful Functions

```{r other-func-demo, echo=TRUE}
# Create vectors
x <- c(5, 10, 15, 20, 25, 30)
names <- c("Alice", "Bob", "Charlie", "David", "Eva")

# Length
length(x)     # Number of elements
length(names)

# Sorting
sort(x)                # Ascending order
sort(x, decreasing = TRUE)  # Descending order
sort(names)            # Alphabetical

# Reversing
rev(x)
rev(names)

# Unique values
y <- c(1, 2, 2, 3, 3, 3, 4, 5, 5)
unique(y)
```

### Exercise 4: Working with Vector Functions

Using this vector of exam scores:
```
scores <- c(85, 92, 78, 65, 88, 75, 96, 82, 78, 90)
```

Calculate and store:
1. The number of students (`num_students`)
2. The average score (`avg_score`)
3. The highest and lowest scores (`score_range`)
4. The sorted scores in descending order (`sorted_scores`)
5. The number of unique scores (`num_unique`)

```{r ex4, exercise=TRUE}
# Create the scores vector
scores <- c(85, 92, 78, 65, 88, 75, 96, 82, 78, 90)

# Write your code here

```

```{r ex4-solution}
# Create the scores vector
scores <- c(85, 92, 78, 65, 88, 75, 96, 82, 78, 90)

# Number of students
num_students <- length(scores)

# Average score
avg_score <- mean(scores)

# Highest and lowest scores
score_range <- range(scores)

# Sorted scores (descending)
sorted_scores <- sort(scores, decreasing = TRUE)

# Number of unique scores
num_unique <- length(unique(scores))

# Display results
num_students
avg_score
score_range
sorted_scores
num_unique
```

```{r ex4-check}
grade_this({
  # Check for pattern matches in the code
  has_num_students <- grepl("num_students\\s*<-\\s*length\\s*\\(\\s*scores\\s*\\)", .user_code)
  
  has_avg_score <- grepl("avg_score\\s*<-\\s*mean\\s*\\(\\s*scores\\s*\\)", .user_code)
  
  has_score_range <- grepl("score_range\\s*<-\\s*range\\s*\\(\\s*scores\\s*\\)", .user_code)
  
  has_sorted_scores <- grepl("sorted_scores\\s*<-\\s*sort\\s*\\(\\s*scores\\s*,\\s*decreasing\\s*=\\s*(TRUE|T)\\s*\\)", .user_code)
  
  has_num_unique <- grepl("num_unique\\s*<-\\s*length\\s*\\(\\s*unique\\s*\\(\\s*scores\\s*\\)\\s*\\)", .user_code)
  
  if (exists("num_students", envir = environment()) && 
      exists("avg_score", envir = environment()) && 
      exists("score_range", envir = environment()) && 
      exists("sorted_scores", envir = environment()) && 
      exists("num_unique", envir = environment())) {
    
    students_match <- num_students == 10
    avg_match <- abs(avg_score - mean(scores)) < 0.0001
    range_match <- all(score_range == range(scores))
    sorted_match <- all(sorted_scores == sort(scores, decreasing = TRUE))
    unique_match <- num_unique == length(unique(scores))
    
    if (students_match && avg_match && range_match && sorted_match && unique_match) {
      pass("Excellent! All your vector function calculations are correct.")
    } else {
      if (!students_match) {
        fail("Your num_students should equal the number of elements in scores (10).")
      } else if (!avg_match) {
        fail("Your avg_score should be the mean of the scores vector.")
      } else if (!range_match) {
        fail("Your score_range should contain the minimum and maximum values of scores.")
      } else if (!sorted_match) {
        fail("Your sorted_scores should be the scores sorted in descending order.")
      } else if (!unique_match) {
        fail("Your num_unique should be the number of unique values in scores.")
      }
    }
  } else if (has_num_students && has_avg_score && has_score_range && has_sorted_scores && has_num_unique) {
    pass("Excellent! All your vector function calculations are correct.")
  } else if (!has_num_students) {
    fail("Make sure to calculate the number of students as num_students <- length(scores).")
  } else if (!has_avg_score) {
    fail("Make sure to calculate the average score as avg_score <- mean(scores).")
  } else if (!has_score_range) {
    fail("Make sure to find the range of scores as score_range <- range(scores).")
  } else if (!has_sorted_scores) {
    fail("Make sure to sort the scores in descending order as sorted_scores <- sort(scores, decreasing = TRUE).")
  } else if (!has_num_unique) {
    fail("Make sure to count unique scores as num_unique <- length(unique(scores)).")
  } else {
    fail("Make sure to complete all five vector function calculations.")
  }
})
```

## Named Vectors

You can assign names to elements in a vector, which can make your code more readable and help with accessing elements by name.

```{r named-vector-demo, echo=TRUE}
# Create a named vector
monthly_sales <- c(Jan = 1500, Feb = 1800, Mar = 2200, Apr = 1900)
monthly_sales

# Alternative way to assign names
sales <- c(1500, 1800, 2200, 1900)
names(sales) <- c("Jan", "Feb", "Mar", "Apr")
sales

# Access elements by name
monthly_sales["Feb"]
monthly_sales[c("Jan", "Mar")]

# Get the names
names(monthly_sales)
```

### Exercise 5: Named Vectors

Create a named vector called `capital_cities` with country names as the names and their capitals as the values for:
- United States (Washington)
- France (Paris)
- Japan (Tokyo)
- Brazil (Brasilia)
- Egypt (Cairo)

Then write code to:
1. Access the capital of Japan
2. Access the capitals of France and Brazil
3. Create a new vector `population` with approximate populations (in millions) for each capital city
4. Make sure the names match those in `capital_cities`

```{r ex5, exercise=TRUE}
# Write your code here

```

```{r ex5-solution}
# Create named vector of capitals
capital_cities <- c("United States" = "Washington", 
                    "France" = "Paris", 
                    "Japan" = "Tokyo", 
                    "Brazil" = "Brasilia",
                    "Egypt" = "Cairo")

# Access Japan's capital
japan_capital <- capital_cities["Japan"]

# Access France and Brazil capitals
france_brazil_capitals <- capital_cities[c("France", "Brazil")]

# Create population vector (approximate values in millions)
population <- c("United States" = 0.7, 
                "France" = 2.2, 
                "Japan" = 14.0, 
                "Brazil" = 3.1,
                "Egypt" = 10.0)

# Display results
capital_cities
japan_capital
france_brazil_capitals
population
```

```{r ex5-check}
grade_this({
  # Check for pattern matches in the code
  has_capital_cities <- grepl('capital_cities\\s*<-\\s*c\\s*\\(', .user_code) && 
                        grepl('["\']United\\s*States["\']\\s*=\\s*["\']Washington["\']', .user_code) &&
                        grepl('["\']France["\']\\s*=\\s*["\']Paris["\']', .user_code) &&
                        grepl('["\']Japan["\']\\s*=\\s*["\']Tokyo["\']', .user_code) &&
                        grepl('["\']Brazil["\']\\s*=\\s*["\']Brasilia["\']', .user_code) &&
                        grepl('["\']Egypt["\']\\s*=\\s*["\']Cairo["\']', .user_code)
  
  has_japan_capital <- grepl('japan_capital\\s*<-\\s*capital_cities\\s*\\[\\s*["\']Japan["\']\\s*\\]', .user_code)
  
  has_france_brazil <- grepl('france_brazil_capitals\\s*<-\\s*capital_cities\\s*\\[\\s*c\\s*\\(\\s*["\']France["\']\\s*,\\s*["\']Brazil["\']\\s*\\)\\s*\\]', .user_code)
  
  has_population <- grepl('population\\s*<-\\s*c\\s*\\(', .user_code) && 
                    grepl('["\']United\\s*States["\']\\s*=', .user_code) &&
                    grepl('["\']France["\']\\s*=', .user_code) &&
                    grepl('["\']Japan["\']\\s*=', .user_code) &&
                    grepl('["\']Brazil["\']\\s*=', .user_code) &&
                    grepl('["\']Egypt["\']\\s*=', .user_code)
  
  if (exists("capital_cities", envir = environment()) && exists("japan_capital", envir = environment()) && 
      exists("france_brazil_capitals", envir = environment()) && exists("population", envir = environment())) {
    
    cities_ok <- is.character(capital_cities) && length(capital_cities) == 5 && 
                all(c("United States", "France", "Japan", "Brazil", "Egypt") %in% names(capital_cities))
    
    japan_ok <- identical(japan_capital, capital_cities["Japan"])
    
    fb_ok <- identical(france_brazil_capitals, capital_cities[c("France", "Brazil")])
    
    pop_ok <- is.numeric(population) && length(population) == 5 && 
             identical(sort(names(population)), sort(names(capital_cities)))
    
    if (cities_ok && japan_ok && fb_ok && pop_ok) {
      pass("Excellent! You've successfully created and worked with named vectors.")
    } else if (!cities_ok) {
      fail("Your capital_cities vector should be a character vector with names for all five countries.")
    } else if (!japan_ok) {
      fail("Your japan_capital should contain just the capital of Japan from the capital_cities vector.")
    } else if (!fb_ok) {
      fail("Your france_brazil_capitals should contain the capitals of France and Brazil from the capital_cities vector.")
    } else if (!pop_ok) {
      fail("Your population vector should be numeric with the same country names as capital_cities.")
    }
  } else if (has_capital_cities && has_japan_capital && has_france_brazil && has_population) {
    pass("Excellent! You've successfully created and worked with named vectors.")
  } else if (!has_capital_cities) {
    fail("Make sure to create a named vector capital_cities with all five countries and their capitals.")
  } else if (!has_japan_capital) {
    fail("Make sure to access Japan's capital with japan_capital <- capital_cities[\"Japan\"].")
  } else if (!has_france_brazil) {
    fail("Make sure to access France and Brazil capitals with france_brazil_capitals <- capital_cities[c(\"France\", \"Brazil\")].")
  } else if (!has_population) {
    fail("Make sure to create a population vector with the same country names as capital_cities.")
  } else {
    fail("Make sure to complete all parts of the exercise.")
  }
})
```

## Summary

In this module, you've learned about vectors in R, including:

1. How to create vectors using `c()`, sequence functions, and repetition
2. Performing arithmetic operations on vectors
3. Vector recycling when operating on vectors of different lengths
4. Accessing vector elements using position, negative, and logical indexing
5. Using common vector functions for statistical operations and manipulation
6. Creating and working with named vectors

Vectors are the foundation for many other data structures in R, and understanding them is crucial for effective R programming.

Next, we'll explore matrices, which extend the concept of vectors to two dimensions.
